---
title: "County Level Texas Controlled Substance Prescription Data"
author: "Julia Silge"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE,
                      warning = FALSE, message = FALSE, 
                      dpi = 180)
options(width=80, tigris_use_cache = TRUE)
library(tidyverse)
library(broom)
library(silgelib)
library(scales)
theme_set(theme_plex())
#theme_set(theme_minimal())
```


## Clean and tidy controlled substance utilization data

Let's open up the dataset and start munging and preparing it.

```{r opioids_raw}
library(tidyverse)
library(readxl)
library(lubridate)
library(googlesheets)

path <- "CountyDrugPillQty_2017_07.xlsx"

opioids_raw <- path %>%
  excel_sheets() %>%
  set_names() %>%
  map_df(~ read_excel(path = path, sheet = .x), .id = "sheet") %>%
  mutate(Date = dmy(str_c("01-", sheet))) %>%
  select(-sheet) %>%
  rename(Name = `Generic Name`)

new_opioids_sheet <- gs_title("TX CS Qty By Drug Name-County")

new_opioids_raw <- new_opioids_sheet %>%
  gs_read("TX CS RX By Generic Name-County",
          col_types = "cnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnn",
          skip = 4,
          verbose = FALSE) %>%
  rename(Name = `Date/Month Filter`)  %>% 
  mutate(Date = case_when(str_detect(Name, 
                                     "^[a-zA-Z]{3}-[0-9]{2}$") ~ Name,
                          TRUE ~ NA_character_)) %>%
  fill(Date, .direction = "down") %>%
  select(-`Grand Total`) %>%
  filter(Name != Date) %>%
  mutate(Date = dmy(str_c("01-", Date))) %>%
  select(Name, Date, everything())
```

We have overlapping measurements for the same drugs and counties from February to June of 2017. How do these measurements compare to each other?

```{r compare, dependson="opioids_raw", fig.width=8, fig.height=5}
## comparing the counts for the overlapping time period

compare <- opioids_raw %>%
  filter(Date > "2017-01-01") %>%
  gather(County, PillsOld, ANDERSON:ZAVALA) %>%
  inner_join(new_opioids_raw %>% 
               gather(County, PillsNew, ANDERSON:ZAVALA)) %>%
  mutate(PercentChange = abs(PillsOld - PillsNew) / PillsOld)

compare %>%    
  ggplot(aes(PillsOld, PillsNew, color = County)) +
  geom_abline(color = "gray50", lty = 2, alpha = 0.8, size = 1.6) +
  geom_point(alpha = 0.4, size = 1.7, show.legend = FALSE) +
  labs(x = "Count from old data",
       y = "Count from new data",
       title = "How close are the prescription quantities in each dataset?",
       subtitle = "Comparing the same drug in the same county")

compare %>% 
  filter(PercentChange != Inf,
         PercentChange < 0.9,
         PillsOld < 5e5) %>% 
  summarise(PercentLessThan10 = mean(PercentChange < 0.15),
            PercentChange = median(PercentChange))

```

Most measurements are close, and the median change is about 20%. The new data is modestly higher in prescription quantity. When we have, it we'll use the new values.

```{r opioids_tidy, dependson="opioids_raw"}
categories_sheet <- gs_title("Drug categories")
drug_categories <- categories_sheet %>%
  gs_read("Sheet1", verbose = FALSE) %>%
  rename(Name = `Generic Name`) %>%
  bind_rows(categories_sheet %>%
              gs_read("Sheet2", verbose = FALSE) %>%
              rename(Name = `Generic Name`)) %>%
  filter(Schedule %in% c("II", "III", "IV", "V"))

opioids_tidy <- opioids_raw %>%
  gather(County, PillsOld, ANDERSON:ZAVALA) %>%
  full_join(new_opioids_raw %>% 
              gather(County, PillsNew, ANDERSON:ZAVALA),
            by = c("Name", "Date", "County")) %>%
  mutate(Pills = coalesce(PillsNew, PillsOld),
         Pills = ifelse(Pills > 1e10, NA, Pills)) %>%
  replace_na(replace = list(Pills = 0)) %>%
  mutate(County = str_to_title(County)) %>% 
  select(-PillsNew, -PillsOld) %>%
  left_join(drug_categories, by = "Name") %>%
  select(County, Date, Name, Category, Schedule, Pills) %>%
  filter(Name != "Unspecified",
         !is.na(Schedule)) %>%
  filter(Date < "2018-05-01")

opioids_tidy
```

In this step, we removed the very small number of prescriptions that were missing drug and schedule information ("unspecified"). Now it's ready to go!

## Total number of pills prescribed over time

What does the overall pattern of pills prescribed looked like?

```{r total, dependson="opioids_tidy", fig.width=7, fig.height=5, fig.cap="Controlled substance prescriptions in Texas"}
opioids_tidy %>%
  group_by(Date) %>%
  summarise(Pills = sum(Pills) / 1e6) %>%
  ggplot(aes(Date, Pills)) +
  geom_smooth(method = "lm") +
  geom_line(size = 1.5, alpha = 0.7) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "Pills prescribed per month (million)",
       title = "Controlled substance prescriptions in Texas",
       subtitle = "The median number of pills prescribed per month in Texas during this time period is 200 million")
```

We see evidence for modest decrease over this time period. We need to compare to the population growth in the same time period to make a meaningful statement about the rate.


```{r growth_rate, dependson="opioids_tidy"}
library(broom)

fit_growth <- opioids_tidy %>%
  group_by(Date) %>%
  summarise(Pills = sum(Pills)) %>%
  lm(Pills ~ Date, data = .)

growth_rate <- (tidy(fit_growth) %>% 
                  filter(term == "Date") %>% 
                  pull(estimate)) / (opioids_tidy %>% 
                                       group_by(Date) %>% 
                                       summarise(Pills = sum(Pills)) %>% 
                                       pull(Pills) %>%
                                       median())
```

The number of pills prescribed per month is changing at about `r percent(growth_rate)` each month, or about `r percent(12 * growth_rate)` each year. This is lower than the rate of Texas' population growth, estimated by the [US Census Bureau at about 1.4% annually](https://www.census.gov/newsroom/press-releases/2017/estimates-idaho.html). Given what we find out further below about the racial/ethnic implications of population level opioid use in Texas *and* what groups are driving population growth in Texas, this likely makes sense.

## Which drugs are growing or shrinking the fastest?

Let's examine how these prescriptions are changing with time. Let's use linear regression modeling to find the individual drugs that are being prescribed more often now compared to two years ago. Let's start with the schedules (II, III, IV, V).

```{r schedule, dependson="opioids_tidy", fig.width=8, fig.height=5, fig.cap="Controlled substance prescriptions by schedule"}
opioids_tidy %>%
  count(Schedule, wt = Pills, sort = TRUE) %>%
  mutate(Percent = percent(n / sum(n))) %>%
  select(-n) %>%
  kable(col.names = c("Schedule", "% of total pills over this time period"))

opioids_tidy %>%
  count(Schedule, Date, wt = Pills) %>%
  mutate(Schedule = factor(Schedule, levels = c("II", "III", "IV", "V",
                                                "Unspecified"))) %>%
  ggplot(aes(Date, n, color = Schedule)) +
  geom_line(alpha = 0.8, size = 1.5) +
  expand_limits(y = 0) +
  labs(x = NULL, y = "Pills prescribed per month",
       title = "Controlled substance prescriptions by schedule",
       subtitle = "Schedule IV drugs account for the most doses, with Schedule II close behind")
```

Schedule IV drugs accounted for the most doses prescribed, with Schedule II close behind, and both look like they are decreasing in prescribing prevalence. Let's fit some models.

```{r schedule_models, dependson="opioids_tidy"}
library(broom)

schedule_by_month <- opioids_tidy %>%
  group_by(Schedule, Date) %>%
  summarise(Pills = sum(Pills))

time_models <- schedule_by_month %>%    
  nest(-Schedule) %>%
  mutate(models = map(data, ~ lm(Pills ~ Date, .))) %>%
  unnest(map(models, tidy)) %>%
  filter(term == "Date") %>%
  arrange(desc(estimate))

time_models %>%
  kable()
```

These models are all asking the question, "What is the rate of change of doses prescribed with time?" The p-values are almost all low (several < 0.005), indicating that we are seeing decreases for Schedule IV and II, just like we see in the plot, starting at the beginning of 2017.

Now let's look at specific drug categories like opioid, stimulant, sedative, and so forth. What are the top 10?

```{r}
opioids_tidy %>%
  count(Category, wt = Pills, sort = TRUE) %>%
  mutate(Percent = percent(n / sum(n))) %>%
  top_n(10, n) %>%
  select(-n) %>%
  kable(col.names = c("Drug", "% of total pills over this time period"))

```

Opioids are by far dominant in these controlled substance prescriptions. Let's fit some models, including adjusting the p-value for multiple comparisons.

```{r time_models, dependson="opioids_tidy"}
category_by_month <- opioids_tidy %>%
  group_by(Category, Date) %>%
  summarise(Pills = sum(Pills))

time_models <- category_by_month %>%    
  nest(-Category) %>%
  mutate(models = map(data, ~ lm(Pills ~ Date, .))) %>%
  unnest(map(models, tidy)) %>%
  filter(term == "Date") %>%
  mutate(p.value = p.adjust(p.value, "bonferroni")) %>%
  arrange(desc(estimate))

time_models %>%
  kable()
```

Many of the p-values are high, indicating that they are not being prescribed more or less, but at mostly the same rate over this time period.

Which drugs are being prescribed *more* or *less*? These are the drugs that are show evidence of growing or shrinking over this time period, at the p < 0.05 level.

```{r fastest_growing, dependson="time_models", fig.width=12, fig.height=8, fig.cap="Growing and shrinking controlled substances in Texas"}
opioids_tidy %>%
  inner_join(time_models %>%
               filter(p.value < 0.05), by = "Category") %>%
  count(Date, Category, estimate, wt = Pills) %>%
  ggplot(aes(Date, n, color = estimate > 0)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, lty = 2) +
  scale_y_continuous(labels = scales::comma_format()) +
  expand_limits(y = 0) +
  facet_wrap(~Category, nrow = 3, scales = "free_y") +
  theme(legend.position="none") +
  theme(legend.title=element_blank()) +
  labs(x = NULL, y = "Pills prescribed per month",
       title = "Growing and shrinking controlled substances in Texas",
       subtitle = "Amphetamines are being prescribed more, and barbiturates are being prescribed less")
```

## Where is controlled substance use changing?

Let's use linear modeling to find counties where controlled substance use is changing. We would need to find the rate of population growth for these countries to make strong statements about what these mean.

```{r county_models, dependson="opioids_tidy"}
library(broom)

opioids_by_county <- opioids_tidy %>%
  group_by(County, Date) %>%
  summarise(Pills = sum(Pills))

county_models <- opioids_by_county %>%    
  nest(-County) %>%
  mutate(models = map(data, ~ lm(Pills ~ Date, .))) %>%
  unnest(map(models, tidy)) %>%
  filter(term == "Date") %>%
  arrange(desc(estimate))

county_models    
```

Which counties have seen the biggest increases?

```{r growing_counties, dependson="county_models", fig.width=8, fig.height=6, fig.cap="Counties experiencing increases in controlled substance prescriptions"}
opioids_by_county %>%
  inner_join(county_models %>%
               top_n(5, estimate)) %>%
  ggplot(aes(Date, Pills, color = County)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, lty = 2) +
  scale_y_continuous(labels = scales::comma_format()) +
  expand_limits(y = 0) +
  theme(legend.title=element_blank()) +
  labs(x = NULL, y = "Pills prescribed per month",
       title = "Controlled substance prescriptions in Texas",
       subtitle = "Lubbock and Smith counties are seeing the fastest growth")
```

Smith county is the locatin of Tyler, TX.

What are counties that are experiencing decreases?

```{r shrinking_counties, dependson="county_models", fig.width=8, fig.height=6, fig.cap="Counties not experiencing increases in controlled substance prescriptions"}
opioids_by_county %>%
  inner_join(county_models %>%
               top_n(-5, estimate)) %>%
  ggplot(aes(Date, Pills, color = County)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, lty = 2) +
  scale_y_continuous(labels = scales::comma_format()) +
  expand_limits(y = 0) +
  theme(legend.title=element_blank()) +
  labs(x = NULL, y = "Pills prescribed per month",
       title = "Controlled substance prescriptions in Texas",
       subtitle = "Harris and Collin counties are seeing decreases (or near decreases)")
```

Collin County is a populous county that is part of the DFW area.

## Connecting to Census data

Let's download some Census data to connect to this controlled substance prescription data.

```{r}
library(tidycensus)

population <- get_acs(geography = "county", 
                      variables = "B01003_001", 
                      state = "TX",
                      geometry = TRUE) 

population

household_income <- get_acs(geography = "county", 
                            variables = "B19013_001", 
                            state = "TX",
                            geometry = TRUE) 

household_income
```

To look at geographical patterns, we will take the median number of pills prescribed per month for each county during the time we have data for.

```{r}
opioids_joined <- opioids_by_county %>% 
  ungroup %>%
  mutate(Date = case_when(Date > "2017-01-01" ~ "2017 and later",
                          TRUE ~ "Before 2017")) %>%
  group_by(County, Date) %>% 
  summarise(Pills = median(Pills)) %>% 
  ungroup %>%
  mutate(County = str_to_lower(str_c(County, " County, Texas")),
         County = ifelse(County == "de witt county, texas",
                         "dewitt county, texas", County)) %>%
  inner_join(population %>% mutate(County = str_to_lower(NAME)), by = "County") %>%
  mutate(OpioidRate = Pills / estimate)

```


#### Controlled substance prescription rate in the top 10 most populous Texas counties

```{r dependson="texas_opioids"}
opioids_joined %>% 
  filter(Date == "2017 and later") %>%
  top_n(10, estimate) %>%
  arrange(desc(estimate)) %>%
  select(NAME, OpioidRate) %>%
  kable(col.names = c("County", "Median monthly pills per capita"), digits = 2)
```

These rates vary a lot; the controlled substance prescription rate in Tarrant County is almost 40% higher than the rate in Harris County.

We can also map the state as a whole.

```{r make_map, dependson="opioids_joined", fig.width=12, fig.height=5, fig.cap="Mapping controlled substance prescriptions across Texas"}
library(sf)
library(viridis)

opioids_map <- opioids_joined %>%
  mutate(OpioidRate = ifelse(OpioidRate > 16, 16, OpioidRate))

opioids_map %>%
  mutate(Date = factor(Date, levels = c("Before 2017", "2017 and later"))) %>%
  st_as_sf() %>%
  ggplot(aes(fill = OpioidRate, color = OpioidRate)) + 
  geom_sf() + 
  coord_sf() + 
  facet_wrap(~Date) +
  scale_fill_viridis(labels = comma_format()) + 
  scale_color_viridis(guide = FALSE) +
  labs(fill = "Monthly pills\nper capita")

```


There are low rates in the Rio Grande Valley and high rates in north and east Texas. There has been change over time as controlled prescription rates have decreased.

Is there a direct relationship with income? Do we see connections to the financial status of a county?


```{r, fig.width=7, fig.height=5, fig.cap="Income and controlled substance prescriptions by county"}
opioids_joined %>% 
  filter(OpioidRate < 20) %>%
  group_by(GEOID, Population = estimate) %>% 
  summarise(OpioidRate = median(OpioidRate)) %>%
  inner_join(household_income %>%
               as.data.frame() %>%
               select(-geometry, -variable) %>%
               rename(Income = estimate)) %>%
  ggplot(aes(Income, OpioidRate, size = Population)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", show.legend = FALSE) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(x = "Median household income", 
       y = "Median monthly prescribed pills per capita",
       title = "Income and controlled substance prescriptions",
       subtitle = "There is no clear relationship between income and prescriptions")
```


Not really. There is no strong relationship here apparent visually, and this is confirmed by modeling (no significant linear relationship). We *can* notice that all the extremely populous counties are low in the plot, with lower controlled substance rates compared to counties of similar income.

What about race?

```{r texas_race}
race_vars <- c("P0050003", "P0050004", "P0050006", "P0040003")

texas_race <- get_decennial(geography = "county", 
                            variables = race_vars, 
                            state = "TX",
                            summary_var = "P0010001") 

texas_race
```


```{r, fig.width=9, fig.height=8, fig.cap="Race and controlled substance prescriptions by county"}
race_joined <- texas_race %>%
  mutate(PercentPopulation = value / summary_value,
         variable = fct_recode(variable,
                               White = "P0050003",
                               Black = "P0050004",
                               Asian = "P0050006",
                               Hispanic = "P0040003")) %>%
  inner_join(opioids_joined %>%
               filter(OpioidRate < 20) %>%
               group_by(GEOID, Date) %>% 
               summarise(OpioidRate = median(OpioidRate)))

race_joined %>%
  group_by(NAME, variable, GEOID) %>%
  summarise(Population = median(summary_value),
            OpioidRate = median(OpioidRate),
            PercentPopulation = median(PercentPopulation)) %>%
  ggplot(aes(PercentPopulation, OpioidRate, 
             size = Population, color = variable)) +
  geom_point(alpha = 0.4) +
  #geom_smooth(method = "lm", lty = 2) +
  facet_wrap(~variable) +
  #theme(legend.position="none") +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_color_discrete(guide = FALSE) +
  labs(x = "% of county population in that racial/ethnic group",
       y = "Median monthly pills prescribed per capita",
       title = "Race and controlled substance prescriptions",
       subtitle = "The more white a county is, the higher the median monthly pills prescribed there",
       size = "County\npopulation")
```

The more white a county is, the higher the rate of controlled substance prescription there. The more Hispanic a county is, the lower the rate of controlled substance prescription there. Effects with Black and Asian race are not clear in Texas. Remember that we did not see an effect with income, though.

We can build a model to predict the monthly prescriptions with both race, income, date (2017 and later vs. before 2017), and total population to see what has a significant effect when we control for all of these. Using the simple explanatory model below, we find that

- poorer counties have higher rates,
- counties with larger white population have higher rates,
- the rates of controlled substance prescription was higher in the past, and
- the prescription rate is higher in more populous counties.

We also see from this table that when controlling for county population, the effects from low income and white population are more dramatic.


```{r}
library(huxtable)

opioids <- race_joined %>%
  select(GEOID, OpioidRate, TotalPop = summary_value,
         variable, PercentPopulation, Date) %>%
  spread(variable, PercentPopulation) %>%
  left_join(household_income %>% 
              select(GEOID, Income = estimate)) %>%
  select(-geometry, -GEOID) %>%
  mutate(Income = Income / 1e5,
         OpioidRate = OpioidRate, 
         Date = factor(Date, levels = c("Before 2017", "2017 and later")),
         Date = fct_recode(Date, ` 2017 and later` = "2017 and later"))

lm1 <- lm(OpioidRate ~ Income + White, data = opioids)
lm2 <- lm(OpioidRate ~ Income + White + Date, data = opioids)
lm3 <- lm(OpioidRate ~ Income + Date + log(TotalPop), data = opioids)
lm4 <- lm(OpioidRate ~ Income + White + Date + log(TotalPop), data = opioids)

huxreg(lm1, lm2, lm3, lm4)
```


```{r eval=FALSE}
path <- "~/Dropbox/Opioid Utilization/Data/regression.docx"
ft <- huxreg(lm1, lm2, lm3, lm4) %>% as_flextable()
my_doc <- officer::read_docx(path)
my_doc <- flextable::body_add_flextable(my_doc, ft)
print(my_doc, target = path)
```


Model metrics such as adjusted $R^2$ and log likelihood indicate that the model including income, percent white population, date, and total population on a log scale provides the most explanatory power for the opioid rate. Using the proportion of population that is Hispanic gives a model that is about as good; these are basically interchangeable but opposite in effect. Overall, the $R^2$ of these models is not extremely high (the best model has an adjusted $R^2$ of `r glance(lm4) %>% pull(adj.r.squared) %>% round(3)`) because these models are estimating population level characteristics and there is significant county-to-county variation that is not explained by these four predictors alone. The population level trends are statistically significant and with the effect sizes at the levels shown here.

We can more directly explore the factors involved in this explanatory model (income, ethnicity, time) visually. 

```{r income_race, fig.width=10.5, fig.height=6, fig.cap="How white population, income, and controlled substance usage are related"}
race_joined %>%
  filter(variable == "White") %>%
  left_join(household_income %>% 
              as.data.frame() %>% 
              select(GEOID, Income = estimate)) %>%
  filter(!is.na(Income)) %>%
  mutate(Income = ifelse(Income <= median(Income, na.rm = TRUE), 
                         "Low income", "High income"),
         PercentPopulation = cut_width(PercentPopulation, 0.1)) %>%
  group_by(PercentPopulation, Income, Date) %>%
  summarise(OpioidRate = median(OpioidRate)) %>%
  #complete(PercentPopulation, Income) %>%
  mutate(Date = factor(Date, levels = c("Before 2017", "2017 and later"))) %>%
  ggplot(aes(PercentPopulation, OpioidRate, color = Income, group = Income)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_smooth(method = "lm", lty = 2, se = FALSE) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, NA)) +
  scale_x_discrete(labels = paste0(seq(0, 0.9, by = 0.1) * 100, "%")) +
  theme(legend.position = "top") +
  facet_wrap(~Date) +
  labs(x = "% of county population that is white",
       y = "Median monthly pills prescribed per 1k population",
       color = NULL,
       title = "White population, income, and controlled substance usage",
       subtitle = "Before 2017, the more white a county was, the more low income was associated with more controlled substance usage")
```


This plot illustrates the relationship between white population percentage and income, and how that has changed with time. The difference in controlled substance usage between lower and higher income counties (above and below the median in Texas) changes along the spectrum of counties' population that is white.

The first effect to notice here is that the more white a county is, the higher the rate of controlled substance prescriptions. This was true both before 2017 and for 2017 and later, and for both low-income and high-income groups of counties. The second effect, though, is to compare the slopes of the two lines. Before 2017, the slope was shallower for higher income counties (above the median in Texas), but in lower income counties (below the median in Texas), the slope was steeper, i.e., the increase in prescription rate with white percentage was more dramatic. For 2017 and later, there is no longer a noticeable difference between low-income and high-income counties, although the trend with white population percentage remains. 

What did we find? At the population level, controlled substance prescriptions are associated with how white a population is; before 2017, how low the income of that white population is had an impact and prescription rates were higher overall.


