---
title: "County Level Texas Controlled Substance Prescription Data"
author: "Julia Silge"
date: "`r Sys.Date()`"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE,
                      warning = FALSE, message = FALSE, 
                      dpi = 180)
options(width=80, tigris_use_cache = TRUE)
library(ggplot2)
library(silgelib)
library(scales)
theme_set(theme_plex())
```


## Clean and tidy controlled substance utilization data

Let's open up the dataset and start munging and preparing it.

```{r opioids_raw}
library(tidyverse)
library(readxl)
library(lubridate)
library(googlesheets)

path <- "CountyDrugPillQty_2017_07.xlsx"

opioids_raw <- path %>%
    excel_sheets() %>%
    set_names() %>%
    map_df(~ read_excel(path = path, sheet = .x), .id = "sheet") %>%
    mutate(Date = dmy(str_c("01-", sheet))) %>%
    select(-sheet) %>%
    rename(Name = `Generic Name`)

new_opioids_sheet <- gs_title("TX CS Qty By Drug Name-County")

new_opioids_raw <- new_opioids_sheet %>%
    gs_read("TX CS RX By Generic Name-County",
            col_types = "cnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnnn",
            skip = 4,
            verbose = FALSE) %>%
    rename(Name = `Date/Month Filter`)  %>% 
    mutate(Date = case_when(str_detect(Name, 
                                       "^[a-zA-Z]{3}-[0-9]{2}$") ~ Name,
                            TRUE ~ NA_character_)) %>%
    fill(Date, .direction = "down") %>%
    select(-`Grand Total`) %>%
    filter(Name != Date) %>%
    mutate(Date = dmy(str_c("01-", Date))) %>%
    select(Name, Date, everything())
```

We have overlapping measurements for the same drugs and counties from February to June of 2017. How do these measurements compare to each other?

```{r compare, dependson="opioids_raw", fig.width=8, fig.height=5}
## comparing the counts for the overlapping time period

compare <- opioids_raw %>%
    filter(Date > "2017-01-01") %>%
    gather(County, PillsOld, ANDERSON:ZAVALA) %>%
    inner_join(new_opioids_raw %>% 
                  gather(County, PillsNew, ANDERSON:ZAVALA)) %>%
    mutate(PercentChange = abs(PillsOld - PillsNew) / PillsOld)

compare %>%    
    ggplot(aes(PillsOld, PillsNew, color = County)) +
    geom_abline(color = "gray50", lty = 2, alpha = 0.8, size = 1.6) +
    geom_point(alpha = 0.4, size = 1.7, show.legend = FALSE) +
    labs(x = "Count from old data",
         y = "Count from new data",
         title = "How close are the prescription quantities in each dataset?",
         subtitle = "Comparing the same drug in the same county")

```

Most measurements are close, and the median change is about 20%. The new data is modestly higher in prescription quantity.

```{r opioids_tidy, dependson="opioids_raw"}
categories_sheet <- gs_title("Drug categories")
drug_categories <- categories_sheet %>%
    gs_read("Sheet1", verbose = FALSE) %>%
    rename(Name = `Generic Name`) %>%
  bind_rows(categories_sheet %>%
              gs_read("Sheet2", verbose = FALSE) %>%
              rename(Name = `Generic Name`)) %>%
  filter(Schedule %in% c("II", "III", "IV", "V"))

opioids_tidy <- opioids_raw %>%
    gather(County, Pills, ANDERSON:ZAVALA) %>%
    bind_rows(new_opioids_raw %>% 
                  gather(County, Pills, ANDERSON:ZAVALA)) %>%
    mutate(Pills = ifelse(Pills > 1e10, NA, Pills)) %>%
    replace_na(replace = list(Pills = 0)) %>%
    mutate(County = str_to_title(County)) %>%
    left_join(drug_categories, by = "Name") %>%
    select(County, Date, Name, Category, Schedule, Pills) %>%
    filter(Name != "Unspecified",
           !is.na(Schedule))

opioids_tidy
```

In this step, we removed the very small number of prescriptions that were missing drug and schedule information ("unspecified"). Now it's ready to go!

## Total number of pills prescribed over time

What does the overall pattern of pills prescribed looked like?

```{r total, dependson="opioids_tidy", fig.width=7, fig.height=5, fig.cap="Controlled substance prescriptions in Texas"}
opioids_tidy %>%
    group_by(Date) %>%
    summarise(Pills = sum(Pills)) %>%
    ggplot(aes(Date, Pills)) +
    geom_smooth(method = "lm") +
    geom_line(size = 1.5, alpha = 0.7) +
    expand_limits(y = 0) +
    labs(x = NULL, y = "Pills prescribed per month",
         title = "Controlled substance prescriptions in Texas",
         subtitle = "The median number of pills prescribed per month in Texas during this time period is 200 million")
```

We see evidence for modest growth over this time period. We need to compare to the population growth in the same time period to make a meaningful statement about the rate.


```{r}
library(broom)

fit_growth <- opioids_tidy %>%
    group_by(Date) %>%
    summarise(Pills = sum(Pills)) %>%
    lm(Pills ~ Date, data = .)

growth_rate <- (tidy(fit_growth) %>% 
                    filter(term == "Date") %>% 
                    pull(estimate)) / (opioids_tidy %>% 
                                           group_by(Date) %>% 
                                           summarise(Pills = sum(Pills)) %>% 
                                           pull(Pills) %>%
                                           median())
```

The number of pills prescribed per month is growing at about `r percent(growth_rate)` each month, or about `r percent(12 * growth_rate)` each year. This is lower than the rate of Texas' population growth, estimate by the [US Census Bureau at about 1.4% annually](https://www.census.gov/newsroom/press-releases/2017/estimates-idaho.html). Given what we find out further below about the racial/ethnic implications of population level opioid use in Texas *and* what groups are driving population growth in Texas, this likely makes sense.

## Which drugs are growing or shrinking the fastest?

Let's examine how these prescriptions are changing with time. Let's use linear regression modeling to find the individual drugs that are being prescribed more often now compared to two years ago. Let's start with the schedules (II, III, IV, V).

```{r schedule, dependson="opioids_tidy", fig.width=8, fig.height=5, fig.cap="Controlled substance prescriptions by schedule"}
opioids_tidy %>%
    count(Schedule, wt = Pills, sort = TRUE) %>%
    mutate(Percent = percent(n / sum(n))) %>%
    select(-n) %>%
    kable(col.names = c("Schedule", "% of total pills over this time period"))

opioids_tidy %>%
    count(Schedule, Date, wt = Pills) %>%
    mutate(Schedule = factor(Schedule, levels = c("II", "III", "IV", "V",
                                                  "Unspecified"))) %>%
    ggplot(aes(Date, n, color = Schedule)) +
    geom_line(alpha = 0.8, size = 1.5) +
    expand_limits(y = 0) +
    labs(x = NULL, y = "Pills prescribed per month",
         title = "Controlled substance prescriptions by schedule",
         subtitle = "Schedule IV drugs account for the most doses, with Schedule II close behind")
```

This looks pretty flat, but let's fit some models.

```{r schedule_models, dependson="opioids_tidy"}
library(broom)

schedule_by_month <- opioids_tidy %>%
    group_by(Schedule, Date) %>%
    summarise(Pills = sum(Pills))

time_models <- schedule_by_month %>%    
    nest(-Schedule) %>%
    mutate(models = map(data, ~ lm(Pills ~ Date, .))) %>%
    unnest(map(models, tidy)) %>%
    filter(term == "Date") %>%
    arrange(desc(estimate))

time_models %>%
    kable()
```

These models are all asking the question, "What is the rate of change of doses prescribed with time?" The p-values are all high (all > 0.05), indicating that we aren't seeing increases or decreases for any schedule, just like we see in the plot. The overall numbers of pills prescribed per month is mostly flat.

Now let's look at specific drug categories like opioid, stimulant, sedative, and so forth. What are the top 10?

```{r}
opioids_tidy %>%
    count(Category, wt = Pills, sort = TRUE) %>%
    mutate(Percent = percent(n / sum(n))) %>%
    top_n(10, n) %>%
    select(-n) %>%
    kable(col.names = c("Schedule", "% of total pills over this time period"))

```

Opioids are by far dominant in these controlled substance prescriptions. Let's fit some models, including adjusting the p-value for multiple comparisons.

```{r time_models, dependson="opioids_tidy"}
category_by_month <- opioids_tidy %>%
    group_by(Category, Date) %>%
    summarise(Pills = sum(Pills))

time_models <- category_by_month %>%    
    nest(-Category) %>%
    mutate(models = map(data, ~ lm(Pills ~ Date, .))) %>%
    unnest(map(models, tidy)) %>%
    filter(term == "Date") %>%
    mutate(p.value = p.adjust(p.value)) %>%
    arrange(desc(estimate))

time_models %>%
    kable()
```

Many of these p-values are high, indicating that they are not being prescribed more or less, but at mostly the same rate over this time period.

Which drugs are being prescribed *more* or *less*? These are the drugs that are show evidence of growing or shrinking over this time period, at the p < 0.05 level.

```{r fastest_growing, dependson="time_models", fig.width=12, fig.height=8, fig.cap="Growing and shrinking controlled substances in Texas"}
opioids_tidy %>%
    inner_join(time_models %>%
                   filter(p.value < 0.05), by = "Category") %>%
    count(Date, Category, estimate, wt = Pills) %>%
    ggplot(aes(Date, n, color = estimate > 0)) +
    geom_line(size = 1.5, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, lty = 2) +
    scale_y_continuous(labels = scales::comma_format()) +
    expand_limits(y = 0) +
    facet_wrap(~Category, nrow = 3, scales = "free_y") +
    theme(legend.position="none") +
    theme(legend.title=element_blank()) +
    labs(x = NULL, y = "Pills prescribed per month",
         title = "Growing and shrinking controlled substances in Texas",
         subtitle = "Amphetamines are being prescribed more, and barbiturates are being prescribed less")
```

## Where is controlled substance use changing?

Let's use linear modeling to find counties where controlled substance use is changing.

```{r county_models, dependson="opioids_tidy"}
library(broom)

opioids_by_county <- opioids_tidy %>%
    group_by(County, Date) %>%
    summarise(Pills = sum(Pills))

county_models <- opioids_by_county %>%    
    nest(-County) %>%
    mutate(models = map(data, ~ lm(Pills ~ Date, .))) %>%
    unnest(map(models, tidy)) %>%
    filter(term == "Date") %>%
    arrange(desc(estimate))

county_models    
```

Which counties have seen the biggest increases?

```{r growing_counties, dependson="county_models", fig.width=8, fig.height=6, fig.cap="Counties experiencing increases in controlled substance prescriptions"}
opioids_by_county %>%
    inner_join(county_models %>%
                   top_n(5, estimate)) %>%
    ggplot(aes(Date, Pills, color = County)) +
    geom_line(size = 1.5, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, lty = 2) +
    scale_y_continuous(labels = scales::comma_format()) +
    expand_limits(y = 0) +
    theme(legend.title=element_blank()) +
    labs(x = NULL, y = "Pills prescribed per month",
         title = "Controlled substance prescriptions in Texas",
         subtitle = "Dallas, Tarrant, and Denton are seeing the fastest growth")
```


What are counties that are experiencing decreases?

```{r shrinking_counties, dependson="county_models", fig.width=8, fig.height=6, fig.cap="Counties not experiencing increases in controlled substance prescriptions"}
opioids_by_county %>%
    inner_join(county_models %>%
                   top_n(-5, estimate)) %>%
    ggplot(aes(Date, Pills, color = County)) +
    geom_line(size = 1.5, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, lty = 2) +
    scale_y_continuous(labels = scales::comma_format()) +
    expand_limits(y = 0) +
    theme(legend.title=element_blank()) +
    labs(x = NULL, y = "Pills prescribed per month",
         title = "Controlled substance prescriptions in Texas",
         subtitle = "Harris and Travis counties are seeing decreases (or near decreases)")
```



## Connecting to Census data

Let's download some Census data to connect to this controlled substance prescription data.

```{r}
library(tidycensus)

population <- get_acs(geography = "county", 
                      variables = "B01003_001", 
                      state = "TX",
                      geometry = TRUE) 

population

household_income <- get_acs(geography = "county", 
                            variables = "B19013_001", 
                            state = "TX",
                            geometry = TRUE) 

household_income
```

To look at geographical patterns, we will take the median number of pills prescribed per month for each county during the time we have data for.

```{r}
opioids_joined <- opioids_by_county %>% 
    group_by(County) %>% 
    summarise(Pills = median(Pills)) %>% 
    mutate(County = str_to_lower(str_c(County, " County, Texas")),
           County = ifelse(County == "de witt county, texas",
                           "dewitt county, texas", County)) %>%
    inner_join(population %>% mutate(County = str_to_lower(NAME)), by = "County") %>%
    mutate(OpioidRate = Pills / estimate * 1e3)

```


#### Controlled substance prescription rate in the top 10 most populous Texas counties

```{r dependson="texas_opioids"}
opioids_joined %>% 
    top_n(10, estimate) %>%
    arrange(desc(estimate)) %>%
    select(NAME, OpioidRate) %>%
    kable(col.names = c("County", "Median monthly pills per 1k population"))
```

These rates vary a lot; the controlled substance prescription rate in Tarrant County is almost 40% higher than the rate in Harris County.

We can also map the state as a whole.

```{r make_map, dependson="opioids_joined", fig.width=8, fig.height=5, fig.cap="Mapping controlled substance prescriptions across Texas"}
library(sf)
library(viridis)

opioids_map <- opioids_joined %>%
    mutate(OpioidRate = ifelse(OpioidRate > 1.6e4, 1.6e4, OpioidRate))

opioids_map %>%
    st_as_sf() %>%
    ggplot(aes(fill = OpioidRate, color = OpioidRate)) + 
    geom_sf() + 
    coord_sf() + 
    scale_fill_viridis(labels = comma_format()) + 
    scale_color_viridis(guide = FALSE) +
    labs(fill = "Monthly pills\nper 1k population")

```


There are low rates in the Rio Grande Valley and high rates in north and east Texas.

Is there a direct relationship with income? Do we see connections to the financial status of a county?


```{r, fig.width=7, fig.height=5, fig.cap="Income and controlled substance prescriptions by county"}
opioids_joined %>% 
    filter(OpioidRate < 2e4) %>%
    group_by(GEOID, Population = estimate) %>% 
    summarise(OpioidRate = median(OpioidRate)) %>%
    inner_join(household_income %>%
                   as.data.frame() %>%
                   select(-geometry, -variable) %>%
                   rename(Income = estimate)) %>%
    ggplot(aes(Income, OpioidRate, size = Population)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", show.legend = FALSE) +
    scale_x_continuous(labels = scales::dollar_format()) +
    scale_y_continuous(labels = scales::comma_format()) +
    labs(x = "Median household income", 
         y = "Median monthly prescribed pills per 1k population",
         title = "Income and controlled substance prescriptions",
         subtitle = "There is no clear relationship between income and prescriptions")
```


Not really. There is no strong relationship here apparent visually, and this is confirmed by modeling (no significant linear relationship). We *can* notice that all the extremely populous counties are low in the plot, with lower controlled substance rates compared to counties of similar income.

What about race?

```{r texas_race}
race_vars <- c("P0050003", "P0050004", "P0050006", "P0040003")

texas_race <- get_decennial(geography = "county", 
                            variables = race_vars, 
                            state = "TX",
                            summary_var = "P0010001") 

texas_race
```


```{r, fig.width=9, fig.height=8, fig.cap="Race and controlled substance prescriptions by county"}
race_joined <- texas_race %>%
    mutate(PercentPopulation = value / summary_value,
           variable = fct_recode(variable,
                                 White = "P0050003",
                                 Black = "P0050004",
                                 Asian = "P0050006",
                                 Hispanic = "P0040003")) %>%
    inner_join(opioids_joined %>%
                   filter(OpioidRate < 2e4) %>%
                   group_by(GEOID) %>% 
                   summarise(OpioidRate = median(OpioidRate)))

race_joined %>%
    rename(Population = summary_value) %>%
    ggplot(aes(PercentPopulation, OpioidRate, 
               size = Population, color = variable)) +
    geom_point(alpha = 0.4) +
    #geom_smooth(method = "lm", lty = 2, se = FALSE) +
    facet_wrap(~variable) +
    #theme(legend.position="none") +
    scale_x_continuous(labels = scales::percent_format()) +
    scale_y_continuous(labels = scales::comma_format()) +
    scale_color_discrete(guide = FALSE) +
    labs(x = "% of county population in that racial/ethnic group",
         y = "Median monthly pills prescribed per 1k population",
         title = "Race and controlled substance prescriptions",
         subtitle = "The more white a county is, the higher the median monthly pills prescribed there",
         size = "County\npopulation")
```

The more white a county is, the higher the rate of controlled substance prescription there. The more Hispanic a county is, the lower the rate of controlled substance prescription there. Remember that we did not see an effect with income, though.

We can build a model to predict the monthly prescriptions with both race and income and see what has an effect when you control for both. Using the simple exploratory model below, it looks like it may be an interaction of race and income, i.e., poorer counties with large white populations have high rates of controlled substance use.


```{r, fig.width=7, fig.height=5}
model_opioids <- race_joined %>%
    select(GEOID, OpioidRate, TotalPop = summary_value,
           variable, PercentPopulation) %>%
    spread(variable, PercentPopulation) %>%
    left_join(household_income %>% 
                  select(GEOID, Income = estimate)) %>%
    select(-geometry, -GEOID) %>%
    lm(OpioidRate ~ Income * White, data = .)

summary(model_opioids)
```

Including total population in this model does not make it a better model; total population is not a significant predictor and it does not improve the model interms of $R^2$ or residuals. Using the proportion of population that is Hispanic gives a model that is about as good; these are basically interchangeable but opposite in effect.

We can explore the interaction between income and ethnicity visually. 

```{r, fig.width=10, fig.height=8, fig.cap="How white population, income, and controlled substance usage interact"}
race_joined %>%
    filter(variable == "White") %>%
    left_join(household_income %>% 
                  as.data.frame() %>% 
                  select(GEOID, Income = estimate)) %>%
    filter(!is.na(Income)) %>%
    mutate(Income = ifelse(Income <= median(Income, na.rm = TRUE), 
                           "Low income", "High income"),
           PercentPopulation = cut_width(PercentPopulation, 0.1)) %>%
    group_by(PercentPopulation, Income) %>%
    summarise(OpioidRate = median(OpioidRate)) %>%
    complete(PercentPopulation, Income) %>%
    ggplot(aes(PercentPopulation, OpioidRate, color = Income, group = Income)) +
    geom_line(size = 1.5, alpha = 0.8) +
    scale_y_continuous(labels = scales::comma_format(),
                       limits = c(0, NA)) +
    scale_x_discrete(labels = paste0(seq(0, 0.9, by = 0.1) * 100, "%")) +
    labs(x = "% of county population that is white",
         y = "Median monthly pills prescribed per 1k population",
         color = NULL,
         title = "White population, income, and controlled substance usage",
         subtitle = "The more white a county is, the more low income is associated with more controlled substance usage")
```


This plot illustrates the interaction between white population percentage and income. The difference in controlled substance usage between lower and higher income counties (above and below the median in Texas) changes along the spectrum of counties' population that is white.

The first effect to notice here is that the more white a county is, the higher the rate of controlled substance prescriptions, for both groups of counties. The second thing, though, is to compare the slopes of the two lines. In higher income counties (above the median in Texas), the slope is shallower, but in lower income counties (below the median in Texas), the slope is steeper, i.e., the increase in prescription rate with white percentage is more dramatic. At the population level, controlled substance prescriptions are associated with how white a population is, and how low the income of that white population is.


## Connecting to overdoses

We can also use the data gathered on overdoses due to prescriptions controlled substances and other drugs to see what connections there are.

```{r}

path <- "UTSouthwestern170731Opioids.xls"

overdoses_tidy <- path %>%
    excel_sheets() %>%
    set_names() %>%
    map_df(~ read_excel(path = path, sheet = .x, skip = 9), .id = "sheet") %>%
    gather(Date, Overdoses, `201301`:`201706`) %>%
    mutate(Date = ymd(str_c(Date, "01")),
           Overdoses = case_when(Overdoses == "*" ~ "Less than 5",
                                 as.numeric(Overdoses) < 10 ~ "5 to 10",
                                 as.numeric(Overdoses) > 10 ~ "More than 10",
                                 is.na(Overdoses) ~ "Zero"),
           Overdoses = factor(Overdoses, levels = c("Zero", "Less than 5",
                                                    "5 to 10", "More than 10")))
```



```{r, fig.width=8, fig.height=7, fig.cap="Overdoses and prescriptions in Texas counties"}
overdoses_tidy %>%
    inner_join(opioids_by_county %>%
                   ungroup %>%
                   mutate(County = toupper(County)), 
               by = c("County", "Date")) %>%
    filter(!is.na(Overdoses)) %>%
    ggplot(aes(Overdoses, Pills, fill = sheet)) +
    geom_boxplot(outlier.alpha = 0.5, alpha = 0.8,
                 show.legend = FALSE, position = "dodge") +
    facet_wrap(~sheet) +
    scale_y_log10() +
    labs(y = "Median monthly pills prescribed",
         title = "Overdoses and prescriptions in Texas counties",
         subtitle = "Counties with higher rates of prescriptions have higher rates of overdoses from all classes of drugs")
```

Having higher rates of overdoses (more than 10 in a month) is rare but we do see that counties with higher numbers of overdoses in a month are the counties with higher numbers of controlled substance prescriptions. The counties with zero overdoses have a broad range of controlled substance prescription rates.

We see that having higher prescriptions rates in a month is associated in raw number with higher overdoses in a month for *all* types of drugs. We want to know how controlled substance prescriptions affect this, to see if the increase is larger for, say, heroin and opioids than for unrelated drugs like cocaine, and how this changes when controlling for population (large vs. small counties).


```{r, fig.width=9, fig.height=6, fig.cap="Overdoses, prescriptions, and population in Texas counties"}
overdoses_joined <- overdoses_tidy %>% 
    group_by(sheet, County, Overdoses) %>% 
    summarize(Total = n()) %>% 
    filter(!is.na(Overdoses)) %>% 
    group_by(sheet, County) %>% 
    filter(Total == last(Total)) %>%
    ungroup %>%
    left_join(opioids_joined %>%
                  transmute(County = toupper(str_replace_all(County, " county, texas", "")),
                            Population = estimate,
                            Pills) %>%
                  mutate(County = ifelse(County == "DEWITT",
                                         "DE WITT", County)))

model_results <- overdoses_joined %>%
    mutate(Population = scale(Population),
           Pills = scale(Pills)) %>%
    nest(-sheet) %>%
    mutate(models = map(data, ~ ordinal::clm(Overdoses ~ Population + Pills, data = .))) %>%
    unnest(map(models, tidy)) %>%
    filter(term %in% c("Population", "Pills")) %>%
    arrange(sheet) %>%
    select(-statistic, -p.value) 

library(ggrepel)

model_results %>%
    select(-std.error) %>%
    spread(term, estimate) %>%
    left_join(model_results %>%
                  select(-estimate) %>%
                  spread(term, std.error),
              by = "sheet",
              suffix = c("", "Error")) %>%
    ggplot(aes(Population, Pills, label = sheet)) +
    geom_hline(size = 1.5, lty = 2, color = "gray50",
               yintercept = 0, alpha = 0.5) +
    geom_vline(size = 1.5, lty = 2, color = "gray50",
               xintercept = 0, alpha = 0.5) +
    geom_errorbar(aes(ymin = Pills - PillsError,
                      ymax = Pills + PillsError),
                  color = "gray80") +
    geom_errorbarh(aes(xmin = Population - PopulationError,
                       xmax = Population + PopulationError),
                   color = "gray80") +
    geom_text_repel(family = "IBMPlexSans-Medium") +
    geom_point() +
    labs(x = "Effect of increasing population on overdoses (scaled)",
         y = "Effect of increasing monthly prescriptions on overdoses (scaled)",
         title = "Overdoses, prescriptions, and population in Texas counties",
         subtitle = "Methamphetamine and heroin overdoses are associated with higher controlled prescription rates")
```

Overdoses are rare overall so we are limited in how precisely we can measure how they depend on other quantities at a population level. When controlling for population, the effect size on overdoses from increasing monthly controlled substance prescriptions is significant for methamphetamines and heroin but not for any other category of overdoses, including opioids themselves. (Remember that the controlled substance prescriptions are dominated by opioids.) Prescription opioid and benzodiazepine overdoses increase with population, but we do not see evidence in this dataset that they are related to increasing prescriptions.

Let's look at how the overdoses are distributed.

```{r, fig.width=15, fig.height=4, fig.cap="Monthly mean overdoses in Texas counties"}
overdoses_map <- opioids_joined %>%
    select(County) %>%
    left_join(overdoses_tidy %>%
                  mutate(OverdosesNumber = case_when(Overdoses == "Zero" ~ 0,
                                                     Overdoses == "Less than 5" ~ 3,
                                                     Overdoses == "5 to 10" ~ 7,
                                                     Overdoses == "More than 10" ~ 10)) %>%
                  group_by(sheet, County) %>%
                  summarise(Overdoses = mean(OverdosesNumber, na.rm = TRUE)) %>%
                  ungroup %>%
                  mutate(County = if_else(County == "DEWITT", "DE WITT", County),
                         County = str_to_lower(County),
                         County = str_c(County, " county, texas"))) %>%
    complete(sheet, County, fill = list(Overdoses = 0)) %>%
    filter(!is.na(sheet)) %>%
    inner_join(opioids_joined %>% 
                   select(County, GEOID, NAME, geometry))

overdoses_map %>%
    st_as_sf() %>%
    ggplot(aes(fill = Overdoses, color = Overdoses)) + 
    geom_sf() + 
    coord_sf() + 
    facet_wrap(~sheet, nrow = 1) +
    scale_fill_viridis(labels = comma_format()) + 
    scale_color_viridis(guide = FALSE) +
    labs(fill = "Overdoses",
         title = "Monthly mean overdoses in Texas counties",
         subtitle = "Only heroin and methamphetamine exhibit evidence of increasing with controlled substance prescriptions, after controlling for population")

```

These maps show that, naturally, there are more overdoses where there are more people. When fitting a model that controls for both population *and* number of pills prescribed, we start to see the differences described above.
