---
title: "Opioid Utilization Exploratory Data Analysis"
author: "Julia Silge"
date: "3/3/2017"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE,
                      warning = FALSE, message = FALSE, 
                      dpi = 180)
options(width=80)
library(ggplot2)
library(silgelib)
theme_set(theme_roboto())
```


## Clean and tidy opioid utilization data

Let's open up the dataset and start munging and preparing it.

```{r opioids_raw}
library(acs)
library(tidyverse)
library(readxl)
library(stringr)
library(reshape2)

opioids_raw <- bind_rows(
    read_excel("./OpiodDispensation_Zip_2014.xlsx", 
               sheet = 1, skip = 1) %>%
        mutate(month = "September") %>%
        rename(zipcode = `Zip Code`),
    read_excel("./OpiodDispensation_Zip_2014.xlsx", 
               sheet = 2, skip = 1) %>%
        mutate(month = "October") %>%
        rename(zipcode = `Zip Code`),
    read_excel("./OpiodDispensation_Zip_2014.xlsx", 
               sheet = 3, skip = 1) %>%
        mutate(month = "November") %>%
        rename(zipcode = `Zip Code`),
    read_excel("./OpiodDispensation_Zip_2014.xlsx", 
               sheet = 4, skip = 1) %>%
        mutate(month = "December") %>%
        rename(zipcode = `Dispensary Postal Code`)) %>%
    rename(total = Totals) 

opioids_raw[is.na(opioids_raw)] <- 0

opioids_raw
```

Let's connect [those zipcodes to counties](http://www2.census.gov/geo/docs/maps-data/data/rel/zcta_county_rel_10.txt). Zipcodes are great, but a bigger unit like counties will be better for mapping purposes.

```{r texas_opioids, dependson="opioids_raw"}
zcta_county <- read_csv("http://www2.census.gov/geo/docs/maps-data/data/rel/zcta_county_rel_10.txt") %>%
    select(ZCTA5, STATE, COUNTY, GEOID) %>%
    filter(STATE == 48)

texas_opioids <- opioids_raw %>% 
    inner_join(zcta_county, by = c("zipcode" = "ZCTA5"))

texas_opioids
```


## Download population in each Texas county from US Census

Let's find the population in each Texas county so we can do some relative comparisons.

```{r population}
tx_counties <- geo.make(state="TX", county="*")
popfetch <- acs.fetch(geography = tx_counties, 
                      endyear = 2014,
                      span = 5, 
                      table.number = "B01003",
                      col.names = "pretty")

myfips <- acs::geography(popfetch) %>%  
                mutate(FIPS = str_c(str_sub(str_c("000", state),-2),
                                    str_sub(str_c("000", county),-3))) %>%
                select(FIPS)

acs::geography(popfetch) <- cbind(myfips, acs::geography(popfetch))

population <- melt(estimate(popfetch)) %>%
    as_tibble() %>%
    select(county_fips = Var1, total_pop = value)

population
```

## Join Census data to opioid data

Next let's join the data from the Census to the opioid utilization data. We will calculate a *rate* of opioid utilization that is the number of total prescriptions divided by the population (per 10,000 people).

```{r opioids_joined, dependson = c("texas_opioids", "population")}
opioids_joined <- texas_opioids %>%
    group_by(GEOID, month) %>% 
    summarise(total = sum(total)) %>%
    ungroup %>%
    inner_join(population, by = c("GEOID" = "county_fips")) %>%
    mutate(opioid_rate = total / total_pop * 1e3,
           opioid_rate = ifelse(is.infinite(opioid_rate), NA, opioid_rate))

opioids_joined
```

Now let's reshape this dataset for plotting purposes. We want columns with the rates and totals for each month.

```{r opioids_spread, dependson = "opioids_joined"}
rate_spread <- opioids_joined %>%
    select(GEOID, month, opioid_rate) %>%
    spread(month, opioid_rate, fill = 0)

totals_spread <- opioids_joined %>%
    select(GEOID, month, total) %>%
    spread(month, total, fill = 0)

cutoff <- 400

opioids_spread <- rate_spread %>%
    left_join(totals_spread, by = "GEOID", suffix = c("Rate", "Total")) %>%
    rename(December = DecemberRate,
           November = NovemberRate,
           October = OctoberRate,
           September = SeptemberRate) %>%
    replace_na(list(DecemberTotal = 0,
                    NovemberTotal = 0,
                    OctoberTotal = 0,
                    SeptemberTotal = 0)) %>%
    mutate(December = ifelse(December > cutoff, cutoff, December),
           November = ifelse(November > cutoff, cutoff, November),
           October = ifelse(October > cutoff, cutoff, October),
           September = ifelse(September > cutoff, cutoff, September))

opioids_spread
```


## Making a map

I'll use a [geojson file for Texas counties](https://github.com/TNRIS/tx.geojson) to make this map. The map will show the rate of opioid prescriptions (per 1000 population) in each county, and click on a county to see its name and the number (count, *not* rate) of prescriptions there in that month. Rates that are higher than `r cutoff` prescriptions per 1000 population are represented by the same color as `r cutoff`.

```{r map, dependson="opioids_spread"}
library(leaflet)
library(scales)


counties <- geojsonio::geojson_read("./tx_counties.geojson", what = "sp")

counties@data <- counties@data %>%
    mutate(FIPS = as.integer(as.character(FIPS))) %>%
    left_join(opioids_spread, by = c("FIPS" = "GEOID")) 

counties@data[is.na(counties@data)] <- 0
pal <- colorNumeric(palette = "viridis", domain = c(0, cutoff))

leaflet(counties, width = "100%") %>%
    addProviderTiles("CartoDB.Positron") %>%
    addPolygons(
        popup = ~paste0("<b>", COUNTY, "</b><br/>", comma_format()(SeptemberTotal), " prescriptions"),
        stroke = FALSE, fillOpacity = 0.7, smoothFactor = 0.5,
        color = ~pal(September), group = "September"
    ) %>%
    addPolygons(
        popup = ~paste0("<b>", COUNTY, "</b><br/>", comma_format()(OctoberTotal), " prescriptions"),
        stroke = FALSE, fillOpacity = 0.7, smoothFactor = 0.5,
        color = ~pal(October), group = "October"
    ) %>%
    addPolygons(
        popup = ~paste0("<b>", COUNTY, "</b><br/>", comma_format()(NovemberTotal), " prescriptions"), 
        stroke = FALSE, fillOpacity = 0.7, smoothFactor = 0.5,
        color = ~pal(November), group = "November"
    ) %>%
    addPolygons(
        popup = ~paste0("<b>", COUNTY, "</b><br/>", comma_format()(DecemberTotal), " prescriptions"),
        stroke = FALSE, fillOpacity = 0.7, smoothFactor = 0.5,
        color = ~pal(December), group = "December"
    ) %>%
    addLayersControl(
        baseGroups = c("September", "October", "November", "December"),
        options = layersControlOptions(collapsed = FALSE)
    ) %>%
    addLegend("bottomright", pal = pal, values = c(0, cutoff),
              title = "Prescriptions per 1k population",
              opacity = 1
    )
```

A couple of things to note:

- The highest rates of opioid prescription are in rural counties but there is more uncertainty in measuring the rate there because populations are so low.
- The counties with the highest rates are pretty consistent within these months, indicating that this isn't just noise; this is a real effect of some kind.
- Some of the very rural counties with high rates of presciptions are next to counties with *no* prescriptions so people may be driving from one county to another to buy opioids from a pharmacy. This will drive the rate up in the rural counties that have pharmacies. (Most people certainly buy from a pharmacy in their own county, but this is less likely to be true in a rural area.)

In case it is helpful, here is just a population map of Texas.

```{r dependson=c("map", "population")}
counties@data <- counties@data %>%
    left_join(population, by = c("FIPS" = "county_fips")) 

pal <- colorQuantile(palette = "viridis", domain = counties@data$total_pop, n = 10)

leaflet(counties, width = "100%") %>%
    addProviderTiles("CartoDB.Positron") %>%
    addPolygons(
        popup = ~COUNTY,
        stroke = FALSE, fillOpacity = 0.7, smoothFactor = 0.5,
        color = ~pal(total_pop)
    ) %>%
    addLegend("bottomright", pal = pal, values = counties@data$total_pop,
              title = "Population Percentiles",
              opacity = 1
    )

```



